---
title: "Data 621 - Project"
author: "Maruthi Mutnuri and Atlanta Liu"
date: "April 28, 2020"
output: html_notebook
---

```{r}
library(dplyr)
library(survival)
library(survminer)
library(KMsurv)
```

##### Read the dataset
```{r}
ist<- read.csv("IST_corrected.csv")
```

##### Creating subsets for the 4 treatment groups:
```{r}
ist_control<-subset(ist, ist$RXASP=='N' & ist$RXHEP=='N')
ist_aspirin<- subset(ist, ist$RXASP=='Y' & ist$RXHEP=='N')
ist_L_Hep<- subset(ist, ist$RXASP=='N' & ist$RXHEP=='L')
ist_H_Hep<- subset(ist, ist$RXASP=='N' & (ist$RXHEP=='M' | ist$RXHEP=='H'))
```

##### Creating the Group variable:
```{r}
ist_control$Group = 'Control'
ist_aspirin$Group = 'Aspirin'
ist_L_Hep$Group = 'L_Hep'
ist_H_Hep$Group = 'H_Hep'
```

##### Extracting the variables needed:
```{r}
ist_data <- rbind(ist_control, ist_aspirin, ist_L_Hep, ist_H_Hep)
data <- select(ist_data,'Group', 'DIED', 'TD', 'SEX', 'AGE', 'RSBP', 'RATRIAL', 'RVISINF')
```

##### Censoring the data at 180 days (6 months):
```{r}
data$TD[data$TD > 180] = 180
data$DIED[data$TD > 180] = 0
```

##### factorizing the treatment groups
```{r}
data$Group = factor(data$Group, levels = c('Control', 'Aspirin', 'L_Hep', 'H_Hep'))
```

##### Removing the data with no values for RATRIAL variable and factorizing it:
```{r}
length(data$Group)
data<-subset(data, data$RATRIAL != '' )
data$RATRIAL = factor(data$RATRIAL, levels = c('N', 'Y'))
length(data$Group)
```

##### Viewing the structure of data:
```{r}
str(data)
```

##### Surminer plot:

```{r fig.height = 10, fig.width = 10, fig.align = "center"}
thefit <- survfit(Surv(TD,DIED)~Group, data=data)
p1<- survminer::ggsurvplot(thefit,censor=T, size =1, data=data, ylim = c(0.75, 1), risk.table = TRUE, risk.table.height=.2, fontsize=5)
print(p1)
```

##### Cox PH model with the treatment groups:
```{r}
coxph.fit_group<-coxph(Surv(TD,DIED)~Group, data=data, ties ="breslow")
summary(coxph.fit_group)
```

##### Visualization of the Cox PH model with the treatment groups:
```{r}
ggforest(coxph.fit_group, data = data)
```

##### Cox PH model with all the variables and interactions:
```{r}
coxph.fit_full<-coxph(Surv(TD,DIED)~Group*AGE+Group*SEX+Group*RSBP+Group*RATRIAL+Group*RVISINF, data=data, ties ="breslow")
summary(coxph.fit_full)
```

##### Cox PH reduced model: non-significant variables and interactions removed:   
```{r}
coxph.fit_red1<-coxph(Surv(TD,DIED)~Group*AGE+SEX+RSBP+RATRIAL+RVISINF, data=data, ties ="breslow")
summary(coxph.fit_red1)
```

##### Anova test: lesser model is significant.
```{r}
anova(coxph.fit_full, coxph.fit_red1)
```

##### Cox PH reduced model: non-significant variables and interactions removed:  
```{r}
coxph.fit_red2<-coxph(Surv(TD,DIED)~Group+AGE+RSBP+RATRIAL+RVISINF, data=data, ties ="breslow")
summary(coxph.fit_red2)
```

##### Anova test: lesser model is significant.
```{r}
anova(coxph.fit_red1, coxph.fit_red2)
```

##### Visualization of final reduced Cox PH model:
```{r}
ggforest(coxph.fit_red2, data = data)
```

##### Testing assumptions:
##### 1. Testing Proportional Hazards Assumption
```{r}
test.ph <- cox.zph(coxph.fit_red2)
test.ph
```

##### Below is the Schoenfeld Residuals plot for testing PH Assumtion: Both Age and Global test fail, so the PH assumption is not met.
```{r fig.height = 15, fig.width = 10, fig.align = "center"}
cox.zph.fit = cox.zph(coxph.fit_red2)
zph<-ggcoxzph(cox.zph.fit, font.main = 12, caption = "Schoenfeld Residuals plot")
zph
```

### Checking the PH assumptions for Age alone: 
```{r fig.height = 5, fig.width = 7, fig.align = "center"}
par(pch = '.')
plot(test.ph[2], ylim= c(-0.2, 0.2))    # a plot for the 3rd variable in the fit> 
chart.plotOptions.line.marker.enabled = FALSE
abline(0,0, col=2)
abline(h= coxph.fit_red2$coef[4], col=3, lwd=2, lty=2)

```
##### coefficent ($\beta$) is changing by time from starting until 16 days, after that it is approximately constant

### To use Survsplit, there should not be any observations are time (TD) = 0.
```{r}
length(data$Group)
data <-subset(data, data$TD != 0)
length(data$Group)
```

```{r}
head(data, 10)
```

```{r}
head(data_tf, 10)
```


### Using survsplit to create 3 stratifications based on the analysis from above plot.
```{r}
data_tf <- survSplit(Surv(TD, DIED) ~ ., data= data, cut=c(4, 16), start="tstart", zero=0, episode= "tgroup", id="id")
head(data_tf, 10)
```

##### Structure of the new dataframe
```{r}
str(data_tf)
```

##### Final Model:
```{r}
coxph.fit_FINAL <- coxph(Surv(tstart, TD, DIED) ~ Group+RSBP+RATRIAL+RVISINF + AGE:strata(tgroup), data=data_tf, ties ="breslow")
summary(coxph.fit_FINAL)
```


##### Testing Assumptions:
##### 1. Testing Proportional Hazards Assumption

```{r}
test.ph <- cox.zph(coxph.fit_FINAL)
test.ph
```

##### Below is the Schoenfeld Residuals plot for testing PH Assumtion: All variables and global test pass the PH assumtion. RATRIAL variable p-values is 0.042 but it is close to 0.05.
```{r fig.height = 15, fig.width = 10, fig.align = "center"}
cox.zph.fit = cox.zph(coxph.fit_FINAL)
zph<-ggcoxzph(cox.zph.fit, font.main = 12, caption = "Schoenfeld Residuals plot")
zph
```

##### 2. Testing influential observations
```{r warning=FALSE}
ggcoxdiagnostics(coxph.fit_FINAL, type = "dfbeta", linear.predictions = FALSE)
```

```{r warning=FALSE}
ggcoxdiagnostics(coxph.fit_FINAL, type = "deviance",
                 linear.predictions = FALSE, ggtheme = theme_bw())
```

##### 3. Non Linearity Assumption:
```{r fig.height = 10, fig.width = 10, fig.align = "center"}
ggcoxfunctional(Surv(TD, DIED) ~ AGE + log(AGE) + sqrt(AGE) , data = data)
```

```{r }
ggcoxfunctional(Surv(TD, DIED) ~ RSBP + log(RSBP) + sqrt(RSBP), data = data)
```

##### APPENDIX# 1: Model with higher order terms based on non-linearity assumption check.
```{r}
coxph.fit_HIGH1 <- coxph(Surv(tstart, TD, DIED) ~ Group+RSBP+ I(RSBP^2)+ I(RSBP^3)+ I(AGE^2) +RATRIAL+RVISINF + AGE, data=data_tf, ties ="breslow")
summary(coxph.fit_HIGH1)
```

```{r}
coxph.fit_HIGH2 <- coxph(Surv(tstart, TD, DIED) ~ Group+RSBP+ I(RSBP^2)+ I(AGE^2) +RATRIAL+RVISINF + AGE, data=data_tf, ties ="breslow")
summary(coxph.fit_HIGH2)
```

```{r}
coxph.fit_HIGH3 <- coxph(Surv(tstart, TD, DIED) ~ Group+RSBP+ I(RSBP^2) +RATRIAL+RVISINF + AGE:strata(tgroup), data=data_tf, ties ="breslow")
summary(coxph.fit_HIGH3)
```


##### APPENDIX# 2:To obtain the survival probability, using the summary of survfit.
```{r}
options(max.print=999999)
thefit <- survfit(Surv(TD,DIED)~Group, data=data)
summary(thefit)
```


   







