---
title: "603 Project"
author: "Greg Cameron,Edwin Aguirre,Maruthi Kumar,Atlanta Liu"
date: "December 3rd, 2019"
output: html_notebook
---


#Reading libraries that are needed for the multiple linear regression analysis
```{r libraries}
library(ggplot2)
library(dplyr)
library(mosaic)
library(tidyverse)
library(olsrr)
library(GGally)
library(lmtest)
library(mctest)
library(car)
```
#Reading files , which provided the dependent and independent variables for our multiple linear regression analysis
```{r reading files}
CO2emission<- read.csv("C:/Users/Edwin Aguirre/Desktop/Data Science/Data 604/project/maddison-data-gdp-per-capita-in-2011us.csv")
GDP<- read.csv("C:/Users/Edwin Aguirre/Desktop/Data Science/Data 604/project/maddison-data-gdp-per-capita-in-2011us.csv")
FossilFuelCon<-read.csv("C:/Users/Edwin Aguirre/Desktop/Data Science/Data 604/project/fossil-fuel-consumption-by-fuel-type.csv")
RenewableEng<-read.csv("C:/Users/Edwin Aguirre/Desktop/Data Science/Data 604/project/modern-renewable-energy-consumption.csv")
PopulationData<-read.csv("C:/Users/Edwin Aguirre/Desktop/Data Science/Data 604/project/population.csv")
#new dataset 
Globalco2concentration<-read.csv("C:/Users/Edwin Aguirre/Desktop/Data Science/Data 604/project/global-co-concentration-ppm.csv")
```

##Data wrangling Stage:\
#Minimum and Maximum Treshold, which we're using in the data wrangling stage of the analysis\ 
```{r}
initial.year='1965'
final.year='2016'
```

#Using the filtering function, which takes as input both the year variable( greater than 1965 years to 2016) and the World variable \

```{r}
GDPperCapita=filter(GDP,Entity=='World',Year>=initial.year)
FossilFuelType=filter(FossilFuelCon,Entity=='World',Year>=initial.year)
Renewabletype=filter(RenewableEng,Entity=='World',Year>=initial.year)
Population1=filter(PopulationData,Entity=='World',Year>=initial.year)
GlobalCo2=filter(Globalco2concentration,Entity=='World',Year>=initial.year)

```
#Applying the filter function which takes the input variable year (when less than 2016). This final output dataframes should have the same number of observations
```{r}
GDPperCapita=filter(GDPperCapita,Year<=final.year)
FossilFuelType=filter(FossilFuelType,Year<=final.year)
Renewabletype=filter(Renewabletype,Year<=final.year)
Population1=filter(Population1,Year<=final.year)
GlobalCo2=filter(GlobalCo2,Year<=final.year)
```
#Removing an additional row from the Population row.
```{r}
Population1=Population1[-c(1),]
```

#Creating a master all dataframe\
```{r}
master_all=GlobalCo2
```


#Combining everything into one master dataframe\
#Adding wanted columns to the master all dataframe. The wanted columns from the 
#fossifueltype renewable energy is 4 to 6
```{r}
master_all=cbind(master_all,FossilFuelType[c(4:6)])
```
#Adding wanted columns to the master all dataframe. The wanted columns from the 
#renewable renewable energy is 4 to 7\
```{r}
master_all=cbind(master_all,Renewabletype[c(4:7)])
```
#Adding one more column which is GDP20, from another dataframe to the master all dataframe
```{r}

master_all=cbind(master_all,Population1[c(4)])
```
#Adding one more column which is population\
```{r}

master_all=cbind(master_all,GDPperCapita[c(4)])
```

#Renaming fourth column,11 and 13 column\
```{r}
names(master_all)[4]<-"CO2ppm"
names(master_all)[11]<-"Other.renewables"
names(master_all)[13]<-"GDP.per.capita"
```

#Performing a multiple linear regression approximation\
```{r}
LINEARMODEL_co2=lm(CO2ppm~Oil..terawatt.hours.+Gas..terawatt.hours.+Coal..terawatt.hours.+Hydropower..terawatt.hours.+Wind..terawatt.hours.+Solar..terawatt.hours.+Year+Other.renewables+Population+GDP.per.capita, data=master_all)
summary(LINEARMODEL_co2)
```
#Checking for Multicollinearity on the main terms:
```{r}
vif(LINEARMODEL_co2)
```
#Therefore, the highest independent value is removed, which is Population.


```{r}
LINEARMODEL_co2.Nopop=lm(CO2ppm~Oil..terawatt.hours.+Gas..terawatt.hours.+Coal..terawatt.hours.+Hydropower..terawatt.hours.+Wind..terawatt.hours.+Solar..terawatt.hours.+Year+Other.renewables+GDP.per.capita, data=master_all)
```

#Checking for VIF

```{r}
vif(LINEARMODEL_co2.Nopop)
```
#Therefore, we remove the Gas independent variable as it has the highest VIF value

#Linear model without Gas independent variable
```{r}
LINEARMODEL_co2.NoGas=lm(CO2ppm~Oil..terawatt.hours.+Coal..terawatt.hours.+Hydropower..terawatt.hours.+Wind..terawatt.hours.+Solar..terawatt.hours.+Year+Other.renewables+GDP.per.capita, data=master_all)
```
#Checking for VIF

```{r}
vif(LINEARMODEL_co2.NoGas)
```
#Therefore, we remove the GDP per capita as the it is the highest value VIF value
```{r}
LINEARMODEL_co2.NoGDP=lm(CO2ppm~Oil..terawatt.hours.+Coal..terawatt.hours.+Hydropower..terawatt.hours.+Wind..terawatt.hours.+Solar..terawatt.hours.+Year+Other.renewables, data=master_all)
```
#Checking for Multicollinearity
```{r}
vif(LINEARMODEL_co2.NoGDP)
```
#Removing Hydropower, as it has the highest vif value

```{r}
LINEARMODEL_co2NoHydro=lm(CO2ppm~Oil..terawatt.hours.+Coal..terawatt.hours.+Wind..terawatt.hours.+Solar..terawatt.hours.+Year+Other.renewables, data=master_all)
```

```{r}
vif(LINEARMODEL_co2NoHydro)
```
#Removing the Wind indepedent variable as it has the highest vif value

```{r}
LINEARMODEL_co2NoWind=lm(CO2ppm~Oil..terawatt.hours.+Coal..terawatt.hours.+Solar..terawatt.hours.+Year+Other.renewables, data=master_all)
```

```{r}
vif(LINEARMODEL_co2NoWind)
```
#Therefore, we remove coal because we want to keep year and other reneweables from other analysis

```{r}
LINEARMODEL_co2NoCoal=lm(CO2ppm~Oil..terawatt.hours.+Solar..terawatt.hours.+Year+Other.renewables, data=master_all)
```

```{r}
vif(LINEARMODEL_co2NoCoal)
```
#Removing the Solar indepedent variable because we want to keep year and other renewables analysis 

```{r}
LINEARMODEL_co2NoSolar=lm(CO2ppm~Oil..terawatt.hours.+Year+Other.renewables, data=master_all)
```

```{r}
vif(LINEARMODEL_co2NoSolar)
```
#Final main effects model
```{r}
LINEARMODEL_co2Finalmaineffects=lm(CO2ppm~Oil..terawatt.hours.+Year+Other.renewables, data=master_all)
```

#Checking for interaction terms on main effects, using the individual coefficient t-test \
#H_o: βi = 0 , where i are the main effects\
#H_a: βi ≠ 0 (i = 1, 2, . . . , p), where i are the main effects\
```{r}
LINEAR_InteractMODEL_co2interaction.terms=lm(CO2ppm~(Oil..terawatt.hours.+Year+Other.renewables)^2,
data=master_all)
summary(LINEAR_InteractMODEL_co2interaction.terms)
```
#From the individual t-test,we can observe a p value of 0.6688, which is greater than α = 0.05. This means we fail to reject the null hypotehsis, therefore we will drop the Year:Other renewables coefficient.\
```{r}
LINEAR_InteractMODEL_co2interaction.terms1=lm(CO2ppm~Oil..terawatt.hours.+Year+Other.renewables+Oil..terawatt.hours.*Year+Oil..terawatt.hours.*Other.renewables,
data=master_all)
summary(LINEAR_InteractMODEL_co2interaction.terms1)
```
#We can observe that every variable (main and interaction) is significant.\

#Checking for higher order terms on the main effects\
```{r}
#option 1: using function ggpairs()
world.data <-data.frame(master_all$CO2ppm,master_all$Oil..terawatt.hours.,master_all$Other.renewables,master_all$Year)
#ggpairs(salarydata)
#LOESS or LOWESS: LOcally WEighted Scatter-plot Smoother
ggpairs(world.data,lower = list(continuous = "smooth_loess",combo ="facethist", discrete = "facetbar", na = "na"))
```
#We can observe graphically, that Oil ,Other and Year can have a higher order term.\
```{r}
LINEAR_InteractMODEL_co2.higherorderterms=lm(CO2ppm~Oil..terawatt.hours.+I(Oil..terawatt.hours.^2)+Year+I(Year^2)+Other.renewables+I(Other.renewables^2), data=master_all)
summary(LINEAR_InteractMODEL_co2.higherorderterms)
```
#It can be observed that there is significance in higher order terms.\

```{r}
LINEAR_InteractMODEL_co2.higherordertermsall=lm(CO2ppm~Oil..terawatt.hours.+I(Oil..terawatt.hours.^2)+Year+I(Year^2)+Other.renewables+I(Other.renewables^2)+Oil..terawatt.hours.*Year+Oil..terawatt.hours.*Other.renewables, data=master_all)
summary(LINEAR_InteractMODEL_co2.higherordertermsall)
```
#Doing an Partial F-test\
#H_o: βp−q+1 = βp−q+2 =. . . = βp = 0, where p is the total number of coefficients\
#Ha:at least one βi ≠ 0\
```{r}
anova(LINEAR_InteractMODEL_co2interaction.terms1,LINEAR_InteractMODEL_co2.higherordertermsall)
```
#The P value from the Partial F test is 0.4091, which is greater than α = 0.05. This means that we fail to reject the null hypothesis, which means that the higher order terms must be dropped(reduced model is valid).Therefore, we will keep two interaction terms, and three main effects.\


#The best fitted model is:\
```{r}
LINEAR_FinalMODEL_2_co2=lm(CO2ppm~Oil..terawatt.hours.+Year+Other.renewables+Oil..terawatt.hours.*Year+Oil..terawatt.hours.*Other.renewables, data=master_all)
summary(LINEAR_FinalMODEL_2_co2)
```


## Checking for all possible asumptions in the multiple linear regression analysis :\

#Checking for linearity assumption:\
#For Residual Plot and checking to see if there are any patterns from the regression residuals\
```{r}
ggplot(LINEAR_FinalMODEL_2_co2, aes(x=.fitted, y=.resid)) +geom_point() + geom_smooth()+ geom_hline(yintercept = 0)+ggtitle("Residual Values vs Fitted Values plot for best fit model")
```
#Therefore, we can see that the linearity assumption is met.That is, no patterns are visually present for the residuals(errors).\

#Checking for Independece Assumption:\

```{r}
ggplot(LINEAR_FinalMODEL_2_co2, aes(x=Year, y=.resid)) +geom_point() + geom_smooth()+ geom_hline(yintercept = 0)+ggtitle("Residual Values vs Time Values plot for best fit model")
```
#It can be observed from the graph that there exist some patterns. This means that the error terms are somehow correlated with each other(dependent).Also, it means that if e_i is a positive value, then it provides information about the sign of the next error term e_i+1.In this case, we can observe the previous statement is only true on some sections of years.Therefore, this means that the independent assumption is not met.\ 



#Checking for normality assumption\
#Producing a histogram plot of residuals of best fit model\
```{r}
ggplot(data=master_all, aes(residuals(LINEAR_FinalMODEL_2_co2))) + 
  geom_histogram(breaks = seq(-1,1,by=0.1), col="red", fill="blue") + 
  labs(title="Histogram for Residuals of best fit model") +
  labs(x="residuals", y="Count")
```
#For Q-Q plot\
```{r}
ggplot(master_all, aes(sample=LINEAR_FinalMODEL_2_co2$residuals)) +ggtitle("Q-Q Plot for Residuals for best fit model")+stat_qq() +stat_qq_line()
```
# Conduction a Shapiro wilk test\
```{r}
shapiro.test(LINEAR_FinalMODEL_2_co2$residuals)
```
#The hypothesis is the following:\
#H_o:the sample data are significantly normally distributed\
#H_A:the sample data are not significantly normally distributed\
#From the Shapiro test, the p value is 0.1329, which is greater than 0.05, this means that we fail to reject the null hypothesis. Therefore, the sample data is significant normally distributed.From the histogram of residuals, it can be observed that the general shape does follow a normal distribution, however there seems to be a gap in between(which is where the outliers are).For the Q-Q plot, it seems like the plot itself follows a normal distribution(with some outliers)\
#Therefore, we can conclude that the normality assumption is met.\ 

#To check for homoscedasticity assumption\
```{r}
bptest(LINEAR_FinalMODEL_2_co2)
```
#The hypthesis are the following:\
#H_o: Heteroscedasticity is not present(homoscedasticity)\
#H_A:Heteroscedasticity is present\
#When conducting a Breusch-Pagan Test , the p value is 0.01983 which is <0.05. This means we reject the null hypothesis, which means the ALTERNATE hypothesis is true. In this context, this means that Heteroscedasticity is present, which means that the variances of the error terms are non-constant.

#To perform a residual plot\


```{r}
ggplot(LINEAR_FinalMODEL_2_co2, aes(x=.fitted, y=.resid)) +geom_point() + geom_smooth()+ geom_hline(yintercept = 0)+ggtitle("Residual Values vs Fitted Values plot for best fit model")
```
#It can be observed from the graph that there is no equal variance in the errors.Therefore, from the graphical representation and the Breusch-Pagan test(which is based on a chi square distribution), we can conclude that the homoscedasticity assumption is not met.\


#To check for influential outliers using the leverage graph\

#To plot Cook's distance\

```{r}
plot(LINEAR_FinalMODEL_2_co2,which=5)
```
#It can be observed from the Residuals vs Leverage graph that no observations are influential(i.e all observations are less than when Cook's distance=0.5)\
#Giving cook's distance graph\

```{r}
plot(LINEAR_FinalMODEL_2_co2,pch=18,col="red",which=c(4))
```
#It can be observed that observation number 10 is 0.15, observation 52 is about 0.13, while observation 25 is about 0.12.\

#Producing a leverage value vs observation graph \
```{r}
lev=hatvalues(LINEAR_FinalMODEL_2_co2)
p = length(coef(LINEAR_FinalMODEL_2_co2))
n = nrow(master_all)
plot(rownames(master_all),lev, main = "Leverage in master_all Dataset", xlab=
"observation",
 ylab = "Leverage Value")
abline(h = 2 *p/n, lty = 1)
abline(h = 3 *p/n, lty = 1)
```
#To obtain all the influential outliers from the dataset \
```{r}
p = length(coef(LINEAR_FinalMODEL_2_co2))
n = nrow(master_all)
outlier = lev[lev>(2*p/n)]
print(outlier) 
```


#Removing outliers, which are influential by using the leverage treshold of 2*p/n:

```{r}
master_all.2p=master_all[-c(2,3,52,53),]
```
#applying the best fitted model on the new master_all.2p dataset
```{r}
LINEAR_FinalMODEL_2_co2.2p=lm(CO2ppm~Oil..terawatt.hours.+Year+Other.renewables+Oil..terawatt.hours.*Year+Oil..terawatt.hours.*Other.renewables, data=master_all.2p)
summary(LINEAR_FinalMODEL_2_co2.2p)
```
#It can be observed that all the same independent variables are still siginificant.\

#For the leverage 3P value\ 
```{r}
lev1=hatvalues(LINEAR_FinalMODEL_2_co2)
p = length(coef(LINEAR_FinalMODEL_2_co2))
n = nrow(master_all)
outlier1 = lev1[lev1>(3*p/n)]
print(outlier1) 
```
#Removing outliers, which are influential by using the leverage treshold of 3*p/n:\

```{r}
master_all.3p=master_all[-c(53),]
```


```{r}
LINEAR_FinalMODEL_2_co23p=lm(CO2ppm~Oil..terawatt.hours.+Year+Other.renewables+Oil..terawatt.hours.*Year+Oil..terawatt.hours.*Other.renewables, data=master_all.3p)
summary(LINEAR_FinalMODEL_2_co23p)
```
#It can be observed that all indepedent variables are still significant.\

#To produce predictions \

```{r}
summary(LINEAR_FinalMODEL_2_co2)
```

#Giving unit conversions that are in units of mboe/d (thousands barrels of oil equivalent per day) to TWHr.\
```{r}
Conversion_of_MBOE_DtoTWHr=365*1.7
million=1000000

```
#Obtained oil values in units of TWHr from tables:\
```{r}
oil_values1970=master_all[6,5]
oil_values1980=master_all[16,5]
oil_values1990=master_all[26,5]
oil_values2000=master_all[36,5]
oil_values2010=master_all[46,5]
oil_values2020=92.3*Conversion_of_MBOE_DtoTWHr
oil_values2030=97.9*Conversion_of_MBOE_DtoTWHr
oil_values2040=100.7*Conversion_of_MBOE_DtoTWHr
```

#Obtaining other renweable values from 1970 to 2010 from our master_all dataset
```{r}
total_other.renewables_1970=master_all[6,11]
total_other.renewables_1980=master_all[16,11]
total_other.renewables_1990=master_all[26,11]
total_other.renewables_2000=master_all[36,11]
total_other.renewables_2010=master_all[46,11]
```


#Got these numbers from dataset\
```{r}
year_rateincreasemarine=0.03
year_rateincreasegeothermal=0.04
year_rateincreasebiofuels=0.06
year_2040.2020=2040-2020
year_2030.2020=2030-2020
```

#Could not find extrapolated values of CO2 however, I could find the growth rate from year 2018 so using our dataset value of 2018\
```{r}
year_2020.2018=2020-2018
year_2030.2018=2030-2018
year_2040.2018=2040-2018
growthrate.2018=2.41
```


#Obtaining values from 2020,2030, and 2040:\
```{r}
#other renweables 2020
marine2020=1
geothermal2020=98
biofuels2020=640
#other renweables 2030
marine2030=marine2020+(year_rateincreasemarine)*year_2030.2020
marine2040=marine2020+(year_rateincreasemarine)*year_2040.2020
geothermal2030=geothermal2020+year_rateincreasegeothermal*year_2030.2020
geothermal2040=geothermal2020+year_rateincreasegeothermal*year_2040.2020
biofuels2030=biofuels2020+year_rateincreasebiofuels*year_2030.2020
biofuels2040=biofuels2020+year_rateincreasebiofuels*year_2040.2020
total_other.renewables_2020=marine2020+geothermal2020+biofuels2020
total_other.renewables_2030=marine2030+geothermal2030+biofuels2030
total_other.renewables_2040=marine2040+geothermal2040+biofuels2040
```
#Predicting C02 to 2020 when year =2020, other renewable=739, and oil=57272.15\

#Obtaining Predictions:\
#Predictions on 1970:\
```{r}
CO2ppm_extrapolation1970=predict(LINEAR_FinalMODEL_2_co2, data.frame(Oil..terawatt.hours.=oil_values1970,Year=1970,Other.renewables=total_other.renewables_1970), type="response")
```

#Predictions on 1980:\

```{r}
CO2ppm_extrapolation1980=predict(LINEAR_FinalMODEL_2_co2, data.frame(Oil..terawatt.hours.=oil_values1980,Year=1980,Other.renewables=total_other.renewables_1980), type="response")
```

#Prediction on 1990:\

```{r}
CO2ppm_extrapolation1990=predict(LINEAR_FinalMODEL_2_co2, data.frame(Oil..terawatt.hours.=oil_values1990,Year=1990,Other.renewables=total_other.renewables_1990), type="response")
```

#Prediction on 2000:\

```{r}
CO2ppm_extrapolation2000=predict(LINEAR_FinalMODEL_2_co2, data.frame(Oil..terawatt.hours.=oil_values2000,Year=2000,Other.renewables=total_other.renewables_2000), type="response")
```

#Prediction on 2010:\

```{r}
CO2ppm_extrapolation2010=predict(LINEAR_FinalMODEL_2_co2, data.frame(Oil..terawatt.hours.=oil_values2010,Year=2010,Other.renewables=total_other.renewables_2010), type="response")
```


```{r}
CO2ppm_extrapolation2020=predict(LINEAR_FinalMODEL_2_co2, data.frame(Oil..terawatt.hours.=oil_values2020,Year=2020,Other.renewables=total_other.renewables_2020), type="response")
CO2ppm_extrapolation2020
```
#The predicted value is 419.4836\

#Predicted  C02 to 2030\

#Predicting C02 to 2020 when year =2030, other renewable=741.6, and oil=60746.95\
```{r}
CO2ppm_otherenewables2030=predict(LINEAR_FinalMODEL_2_co2, data.frame(Oil..terawatt.hours.=oil_values2030,Year=2030,Other.renewables=total_other.renewables_2030), type="response")
CO2ppm_otherenewables2030
```
#The predicted value is 440.98\

#Predicted CO2 to 2040\

#Predicting C02 to 2020 when year =2040, other renewable=741.6, and oil=62484.35\
```{r}
CO2ppm_otherenewables2040=predict(LINEAR_FinalMODEL_2_co2, data.frame(Oil..terawatt.hours.=oil_values2040,Year=2040,Other.renewables=total_other.renewables_2040), type="response")
CO2ppm_otherenewables2040
```
#The predicted value is 460.92\



#To produce graphs that shows our predicted value vs our regular value\

#Creating a dataframe to produce a graph\
```{r,include=False}
year_true=c(1970,1980,1990,2000,2010,2020,2030,2040)
year_true1=c(1970,1980,1990,2000,2010,2016)
CO2extrapolated=c(CO2ppm_extrapolation1970,CO2ppm_extrapolation1980,CO2ppm_extrapolation1990,CO2ppm_extrapolation2000,CO2ppm_extrapolation2010,CO2ppm_extrapolation2020,CO2ppm_otherenewables2030,CO2ppm_otherenewables2040)
CO2actual1=c(master_all[6,4],master_all[16,4],master_all[26,4],master_all[36,4],master_all[46,4],master_all[52,4])
```

```{r,include=FALSE}
predictedframe=data.frame(year_true,CO2extrapolated)
predictedframe1=data.frame(year_true1,CO2actual1)
```


#Finally, we come to the conclusion that, the best fit model is given by:\

\begin{equation*}
\begin{split}
\hat Y_{CO2}  	  &= \beta_{0} + \beta_1 X_{Oil} + \beta_2 X_{Year} + \beta_3 X_{Other} + \beta_4 X_{Oil \times Year} + \beta_5 X_{Oil \times Other} \\
          		  &= -1927.6009 - 0.0202(X_{Oil}) + 1.1461(X_{Year}) +  0.0424(X_{Other}) + 1.0163 \times 10^{-5}(X_{Oil \times Year}) \\
       		  & + 1.0589 \times 10^{-6}(X_{Oil \times Other})
\end{split}
\end{equation*}



#To plot our predicted value based on our multiple linear regression relative to our estimated values\

```{r}
ggplot() + 
  geom_line(data = predictedframe1, aes(x = year_true1, y = CO2actual1, color = " True"),size=3) +
  geom_line(data = predictedframe, aes(x = year_true, y = CO2extrapolated,color = "Predicted"),size=1) +xlab('Year') +ylab('CO2Emissions ppm (parts per million)')+ggtitle("Predicted values using best Fit model relative to estimated values")+ labs(color='Values')

```


```{r}
rm(list = ls(all.names = TRUE))
```